Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-03-01 paper.tex
Beat me to caption fix. Quick fix on input size.
typo

This work is guided by two main goals. (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for neear SOL execution to the greatest extent possible allowed by the underlying hardware and programming model.

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’ 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
\begin{figure}
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
    \caption{\emph{Decoupled Lookback.}}
    \label{fig:decoupled}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/ours.pdf}
    \caption{\emph{Decoupled Fallback (ours).}}
    \label{fig:ours}
  \end{subfigure}
  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max. Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis. The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
  \label{fig:dist}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

>>>>

This work is guided by two main goals. (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for near-SOL execution to the greatest extent possible allowed by the underlying hardware and programming model.

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’ 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
\begin{figure}
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
    \caption{\emph{Decoupled Lookback.}}
    \label{fig:decoupled}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/ours.pdf}
    \caption{\emph{Decoupled Fallback (ours).}}
    \label{fig:ours}
  \end{subfigure}
  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{20}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max. Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis. The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
  \label{fig:dist}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.
